---
title: "Lecture 16:ANCOVA"
date: "`r format(Sys.time(), '%d %B %Y')`"

output:
   html_document:
     theme: flatly
     toc: true
     toc_depth: 4
     number_sections: yes
---    

#BEFORE START - Install RMarkdown

To use this file you have to install RMarkdown!

RMarkdown files have extension: .Rmd

#ANCOVA - Analysis of covariance has one continous response variable and at least one continous and categorical explanatory vairable. 

**IMPORTANT THING to keep in mind is that order of categorical variable matters, because data are not-orthogonal (for details see Crawley: Statistics: an introductory using R). **

EXAMPLE: 
Data file - Grazing.csv
Variables: Root-continous, Seed-continous and Grazing-categorical
Question: Which one of the continuos variables is explanatory and which one is response?
Research question: Does diameter of the top of its rootstock influences weight of the seed produced in grazed or ungrazed plants?
Solution: 
Diameter of the top of rootstock (Root) - explanatory variable (x-axis)
Weight of the seed produced - response variable (y-axis)

6 posible solutions to the problem:
1. Both slopes and intercept are different
2. Different intercepts and same slope
3. Same intercept different slopes
4. No slope and different intercepts
5. One slope and one intercept
6. No slope only one intercept

#HOUSEKEEPING

First lets make some houskeeping before we start:

```{r Housekeeping}
rm(list=ls()) # remove everything currently held in the R memory
graphics.off() # close all open graphics windows 
```

#READ IN YOUR DATA

Enter or read in your data from a file as a CSV (comma separated) file from Excel

REMEMBER: Choose your working directory and put file inside that directory!

##CODE EXPLANATION:
Data file - grazing.csv
header=TRUE - in first row are the variable names (Root, Seed and Grazing)
sep="," - means that values in grazing.csv file are separated with comma (look inside .csv data file in Excel)

```{r setup, include=FALSE}
mydata <- read.table("grazing.csv", header=TRUE, sep=",")
```

#EXPLORE DATA

```{r explore data}
head(mydata) # Return the First Part of an Object -usually first six observations
head(mydata, n = 12L) # option - name of the Object, n= number of lines you want to see (12) from the start of the Object
mydata # Return complete Object
```

#PLOT DATA SCATTERPLOT
```{r}
levels(mydata$Grazing) # gives you all levels of the classification variable within variable Grazing
plot(mydata$Root, mydata$Seed, xlab="Root",ylab="Seed",col=mydata$Grazing, pch=as.numeric(mydata$Grazing)) #xlab-means title for x-axis; y-lab- means title for y-axis, col - means colour for each level of Grazing variable (Grazed is first and first colour is black, while Ungrazed is second and second colur is red; pch- means a type of the symbol to represent each level of Grazed variable)

legend("topleft", pch=c(2,1), col=c("red", "black"), c("Ungrazed", "Grazed"))
```
#FIT THE MOST COMPLICATED MODEL

Our assumption is that both slopes and intercept in Grazed and Ungrazed group differ!

Therefore we use INTERACTION in the model1:

```{r}
model1 <- glm( mydata$Seed ~ mydata$Root * mydata$Grazing )

summary(model1)

```
When we analyse coefficients we see that there is no interaction while P-value for interaction is greater than 0.05!



##PRACTICAL EXPLANATION:

There is no difference in slope between Grazed and Unrazed groups!

#FIT NEXT MODEL

Here we just test if there are differences in intercept between these two groups.

```{r}
model2 <- glm( mydata$Seed ~ mydata$Root + mydata$Grazing )

summary(model2)
coefmod2 <- coef(model2)
```

When we analyse coefficients we see that there is difference between Grazed and Ungrazed group while P-value for interaction is lower than 0.05!


#FIT LINES FOR BOTH MODELS

IMPORTANT: ORDERS MATTERS - Grazing variable has two levels Grazed and Ungrazed which are in alphabetical order!

```{r}
plot(mydata$Root, mydata$Seed, xlab="Root",ylab="Seed",col=mydata$Grazing, pch=as.numeric(mydata$Grazing))

legend("topleft", pch=c(2,1), col=c("red", "black"), c("Ungrazed", "Grazed"))
abline (coefmod2[1],coefmod2[2], col="black", lty=1) #First add the line for the Grazed group
abline (coefmod2[1] + coefmod2[3],coefmod2[2], col="red", lty=2 ) #Then add the line for the Ungrazed group
```


#CHECK RESIDUALS OF ANALYSIS

##Extract the residuals for easy recall in model 2 without interaction


```{r}
rsd <- residuals(model2)
```

# Histogram of the residuals

```{r}
hist(rsd,6)
```


## QQ plot of the residuals to assess how well the residuals compare with an ideal normal distribution

```{r}
qqnorm(rsd)
qqline(rsd,col="red")
```



# Check for a trend in the residuals with the X axis variable
```{r}
plot(mydata$Root,rsd)
abline(0,0,col="blue")
```
Any pattern among the differences between model and observed value indicates a failure of the model to adequately represent some aspect of the relationship within the sampled data.

# Save your data (only if you want)

The "list=" command tells us which variables we want to save
The "file=" option tells us what file to save the data to

save( list=ls(), file="grazing_data.rdata" )