---
title:title: "L11: Generating summary statistics (includes an introduction to writing your own functions)"
date: "`r format(Sys.time(), '%d %B %Y')`"

output:
   html_document:
     theme: flatly
     toc: true
     toc_depth: 4
     number_sections: yes
     
---


How to calculate summary statistics of variables, which also includes a quick explanation of how to write your own functions in R Studio.


# Housekeeping


```{r}
rm(list=ls()) # remove everything currently held in the R memory
```




#Summary statistics

Generate some summary statistics on the Petal Lengths


```{r}
mu <- mean(iris$Petal.Length)    # mean
med <- median(iris$Petal.Length)  # median
variance <- var(iris$Petal.Length)     # variance
stan.dev <- sd(iris$Petal.Length)      # standard deviation
n <- length(iris$Petal.Length) # the number of observations
standard.error <- sd(iris$Petal.Length) / sqrt(length(iris$Petal.Length)) # SE of the mean
maximum <- max(iris$Petal.Length, na.rm=T)
minimum <- min(iris$Petal.Length, na.rm=T)
```


```{r}
min(c(1,5,2,NA,3))
min(c(1,5,2,NA,3), na.rm=T)
```


But... we probably want to calculate these metrics for each of the three species seperately. Use the function aggregate() to collect the data into subsets associated with each species, and apply a given function to the subset. It probably helps if we have a quick graph of the data as a reference point for these summary statistics. A boxplot does this and i refer you to the
associated podcast on boxplots for more information

To remmeber how did the data look like:
```{r}
boxplot(iris$Petal.Length ~ iris$Species)
```

```{r}
mu.sepal.lengths <- aggregate(iris$Petal.Length, by=list(iris$Species), mean)
mu.sepal.lengths
# or
(mu.sepal.lengths <- aggregate(iris$Petal.Length, by=list(iris$Species), mean))
```
If you want just simple statistics with build in functions you can use summary function which calculates: 

```{r}
summary(iris$Petal.Length) # for all Petal lengths

tapply(iris$Petal.Length, iris$Species, summary) # for Petal length by species
IQR(iris$Petal.Length)
```



```{r}
z<-c(1,2,4,5)
z
```


We could repeat this for all the metrics, but instead, we can write our own function that calculates a series of useful summary statistics on the data.

Let's specify our own function
```{r}
summary.stats <- function (x,...) {

# "out" will be a vector with names assigned to each entry.
out <- c( n.obs=length(x),
          min=min(x),
          quantile(x,c(0.25,0.5,0.75)),
          max=max(x),
          mean=mean(x),
          sd=sd(x),
          var=var(x),
          se=sd(x)/sqrt(length(x))
          )
          
return(out)  # this tells the function what to return as "the answer"
} # end of function

```

Let's apply our new function on our vector z.
```{r}
summary.stats(z)
```


Now apply our summary.stats() function to the Petal Length data by Species

```{r}
petal.length.summaries <- aggregate(iris$Petal.Length,by=list(iris$Species),summary.stats)
petal.length.summaries
```

However, take care that you need first to run function and then you can use it for the variables of interest within dataset.

ASSIGNMENT:

1. Add interquantile range IQR into your function called summary.stats1 and perform calculation for Sepal.Length and store values into sepal.length.summaries. 

Solution:


```{r}
summary.stats1 <- function (x,...) {

# "out" will be a vector with names assigned to each entry.
out <- c( n.obs=length(x),
          min=min(x),
          quantile(x,c(0.25,0.5,0.75)),
          max=max(x),
          mean=mean(x),
          sd=sd(x),
          var=var(x),
          se=sd(x)/sqrt(length(x)),
          IQR=IQR(x)
          )
          
return(out)  # this tells the function what to return as "the answer"
}
sepal.length.summaries <- aggregate(iris$Sepal.Length,by=list(iris$Species),summary.stats1)
sepal.length.summaries
```

