---
title:title: "L10: Advanced Boxplots"
date: "`r format(Sys.time(), '%d %B %Y')`"

output:
   html_document:
     theme: flatly
     toc: true
     toc_depth: 4
     number_sections: yes
     
---

Advanced techniques for BoxplotsThis example uses the R included data "iris"


# Housekeeping

Remember that `rm(list=ls())` is not sufficient for a full clean set up of R, and you should use `Ctrl/Cmd-SHIFT-F10` in Rstudio to Restart R cleanly and check that this works at least at the end of your analysis, or periodically during development.

```{r}
# remove (almost) all objects currently held in the R environment
rm(list=ls()) 
```



Enter or read in your data from a file. In this example we will use the in-built dataset "iris" which gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica. 

see ?iris for more information

```{r load_data}
data("iris")
str(iris)

```



# A basic default boxplot

```{r}
par(mfrow=c(1,1))
boxplot( Petal.Length ~ Species, data = iris,
		ylab = "Petal Length (cm)", xlab = "Species")

```




# Customising boxplot

Here are some advanced tricks to customising plots.Remove the frame surrounding the boxplot by suppressing the axes and manually specifying them. NB this is straight forward for most plots, but for some reason the boxplot function is encoded differently.

##Comparison of old and new plot without frame

```{r, fig.height=9}

par(mfrow=c(2,1))

boxplot(  Petal.Length ~ Species, data = iris, 
          ylab = "Petal Length (cm)", xlab = "Species") # old one

boxplot(  Petal.Length ~ Species, data = iris,
          frame.plot = FALSE,  ylab = "Petal Length (cm)", xlab = "Species", 
          las = 1) 
```

##Removing of the axes

```{r}
par(mfrow=c(1,1))
boxplot( Petal.Length ~ Species, data = iris,  axes = FALSE) # axes=F removes axes from the plot
```

## Adding the L box shape
```{r}
boxplot(  Petal.Length ~ Species, data = iris,  axes = FALSE)
box(bty="L") # adding L-type box on your plot
```

## Adding information to the x-axis
```{r}
boxplot(  Petal.Length ~ Species, data = iris, axes = FALSE)
box(bty="L")
axis(1, at = 1:3, c("Iris setosa","Iris versicolor","Iris virginica"),
        font = 3, cex.axis = 1.2, tcl = 0.5) 

```

## Adding information to the y-axis

```{r}
boxplot(  Petal.Length ~ Species, data = iris, ylim = c(0,7), axes = FALSE)
box(bty="L", xaxs = "e")
axis(1, at = 1:3, c("Iris setosa","Iris versicolor","Iris virginica"),
        font=3, cex.axis=1.2, tcl=0.5)
title(xlab="Species", cex.lab=1.4)
axis(2, at=0:7, las=1, tcl=0.5) # las=1 changes direction of the y-axis text
title(ylab="Petal Length (cm)",cex.lab=1.2)
```

## Adding lightgray colour to the boxplot

```{r}
boxplot(  Petal.Length ~ Species, data = iris, 
          ylim = c(0,7), axes = F, col = "lightgray")
box(bty="L")
axis(1,at=1:3,c("Iris setosa","Iris versicolor","Iris virginica"),
        font=3, cex.axis=1.2, tcl=0.5)
title(xlab="Species", cex.lab=1.4)
axis(2, at=0:7, las=1, tcl=0.5) # las=1 changes direction of the y-axis text
title(ylab="Petal Length (cm)",cex.lab=1.2)

```

## Adding colour to the boxplot

```{r}
boxplot(  Petal.Length ~ Species, data = iris, 
          ylim = c(0,7), axes = F, col = rainbow(10))
box(bty="L")
axis(1,at=1:3,c("Iris setosa","Iris versicolor","Iris virginica"),
        font=3, cex.axis=1.2, tcl=0.5)
title(xlab="Species", cex.lab=1.4)
axis(2, at=0:7, las=1, tcl=0.5) # las=1 changes direction of the y-axis text
title(ylab="Petal Length (cm)",cex.lab=1.2)

```


# Saving tiff from your plot

N.B. code chunk not evaluated during knitting of html output file.

```{r, eval = FALSE}
tiff("outfile.tif", compression = "lzw") # produces nice tiff

#tiff(file = "temp.tiff", width = 3200, height = 3200, units = "px", res = 800) # produced large size tiff 
boxplot(  Petal.Length ~ Species, data = iris, 
          ylim=c(0,7), axes=F, col=rainbow(10))
box(bty="L")
axis(1,at=1:3,c("Iris setosa","Iris versicolor","Iris virginica"),
        font=3, cex.axis=1.2, tcl=0.5)
title(xlab="Species", cex.lab=1.4)
axis(2, at=0:7, las=1, tcl=0.5) # las=1 changes direction of the y-axis text
title(ylab="Petal Length (cm)", cex.lab=1.2)

dev.off()

```

