---
title: "L10b: Advanced Boxplots using ggplot2"
date: "`r format(Sys.time(), '%d %B %Y')`"

output:
   html_document:
     theme: flatly
     toc: true
     toc_depth: 4
     number_sections: yes
     
---

Advanced techniques for Boxplots. This example uses the R included data "iris" and plots using the package ggplot2.


# Housekeeping

Remember that `rm(list=ls())` is not sufficient for a full clean set up of R, and you should use `Ctrl/Cmd-SHIFT-F10` in Rstudio to Restart R cleanly and check that this works at least at the end of your analysis, or periodically during development.

```{r}
# remove (almost) all objects currently held in the R environment
rm(list=ls()) 
```


# Alternative Boxplots in ggplot2 package

As always, `ggplot2` does boxplots, and arguably easier / prettier.

```{r}
library(ggplot2)
```



## Basic box plot in ggplot2

```{r}

# load the iris dataset to the global environment
data("iris")

# take a look at the structure of the data
str(iris)

# create a basic ggplot with a boxplot geom on top
p1 <- ggplot(iris, aes(x = Species, y = Petal.Length)) + geom_boxplot()
print(p1)
```

## Rotate the box plot

Flip it 90 degrees.

```{r}
# to plot p1, add a coordinate flip
p2 <- p1 + coord_flip()
print(p2)
```

## Notched box plot

The Notch - displays the a confidence interval around the median which is normally based on the median $\pm 1.58 * \text{IQR} / \sqrt n$.  According to Graphical Methods for Data Analysis (Chambers, 1983) although not a formal test the, if two boxes' notches do not overlap there is ‘strong evidence’ (95% confidence) their medians differ. 

```{r}
# redraw our boxplot, and this time add notches
p3 <- ggplot(iris, aes(x = Species, y = Petal.Length)) + 
  geom_boxplot(notch = TRUE)
print(p3)
```


  
## Change outlier, color, shape and size

You can customise the details of the visualisation.

```{r}
# make some adjustments to the way the outliers are rendered
p4 <- ggplot(iris, aes(x = Species, y = Petal.Length)) + 
  geom_boxplot(outlier.colour = "red", outlier.shape = 8, outlier.size = 4)

print(p4)
```



# Colour Boxplot in ggplot2

```{r}
# Add colour to our boxplots, overlay the raw data jittered a bit for 
# clarity, and adjust the theme to a more usual format for publications.
q <- ggplot(data = iris, aes(x = Species, y = Petal.Length)) + 
  
  geom_boxplot(aes(fill = Species), outlier.color = NA) + 
  
  geom_point(aes(y = Petal.Length, group = Species,
                 fill = Species), 
             position = position_jitter(width = 0.15), 
             shape = 21, 
             stroke = 1.1) + 
  
  scale_fill_manual(values = c("pink", "yellow", "lightblue")) +
  
  guides(fill = guide_legend(title="Legend Title")) + 
  xlab("Species") + 
  ylab("Petal length (cm)") + 
  ggtitle("Petal length in several species of iris") + 
  
  theme_classic() + 
  theme(axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 14),
        legend.text  = element_text(size = 14),
        legend.title = element_text(size = 16))

# print our efforts to screen
print(q)
```

