---
title: "Lecture 8: Errorbar plots"
date: "`r format(Sys.time(), '%d %B %Y')`"

output:
   html_document:
     theme: flatly
     toc: true
     toc_depth: 4
     number_sections: yes
     
---

-----------------------------------------------------------
# Housekeeping

Remove everything currently held in the R memory

```{r}
rm(list=ls())
```

#Errorbar plot

Use an errorbar plot to show the data
```{r}
par(mfrow=c(1,1))
```

Calculate the means of the groups using tapply()

```{r}
mu <- tapply(iris$Petal.Length,iris$Species,mean)
```


Here i write my own function to calculate the 95% confidence intervals, see

[WIKIPEDIA](https://en.wikipedia.org/wiki/Standard_error#Assumptions_and_usage)


```{r}
CI95 <- function (x) {return(1.96*sqrt(var(x)/length(x)))} 
se <- tapply(iris$Petal.Length,iris$Species,CI95)
plot(1:3,mu,xlim=c(0,4),ylim=c(1,7),pch=20, las=1, bty="L",
      xlab="Species", ylab="Petal Length (cm)")
arrows( x0=1:3, x1=1:3, y0=mu+se, y1=mu-se, code=3, length=0.1, angle=90)
```


## Fix the x-axis labels  and tidy up the plot


```{r}
par(mfrow=c(1,1))

```

Calculate the means of the groups

```{r}
mu <- tapply(iris$Petal.Length,iris$Species,mean)
CI95 <- function (x) {return(1.96*sd(x)/sqrt(length(x)))} 
se <- tapply(iris$Petal.Length,iris$Species,CI95)
plot(1:3,mu,xlim=c(0,4),ylim=c(1,7),pch=20, las=1, bty="L", xaxt="n",xlab="Species", ylab="Petal Length (cm)")
axis(1, at=1:3, levels(iris$Species)) 
arrows( x0=1:3, x1=1:3, y0=mu+se, y1=mu-se, code=3, length=0.1, angle=90)
```

##Axis function description

axis(side, at=, labels=, pos=, lty=, col=, las=, tck=, ...)

Side option in axis function - 	an integer indicating the side of the graph to draw the axis (1=bottom, 2=left, 3=top, 4=right)
At - 	a numeric vector indicating where tic marks should be drawn

# Errorbar in ggplot

Do not forget to install ggplot package before executing this code.

```{r}
library(tidyverse) # load package with ggplot
ggplot(iris, aes(Species, Petal.Length)) + 
    stat_summary(fun.y = mean, geom = "point") + 
   coord_cartesian(ylim=c(0, 6)) + labs(y = "Petal length (cm)") +
    stat_summary(fun.data = mean_se, geom = "errorbar", width=0.1, size=0.8, colour="blue") 

```

